---
title: "DABOM Data Prep 'Cheatsheet'"
author: Kevin See
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{2.1) DABOM Data Prep 'Cheatsheet'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
# knitr options
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  eval = FALSE,
  comment = "#>"
)
# turn off one check
options(rmarkdown.html_vignette.check_title = FALSE) 
```


# Introduction

The core of `PITcleanr`'s functionality relies on detections from [PTAGIS](https://ptagis.org/), a configuration file that maps those detections onto user-defined "nodes", and a parent-child table showing how those nodes are related to each other on a stream network (i.e. which nodes are connected). `PITcleanr` contains several functions to help create the configuration file and the parent-child table, but the user can also create those files by hand, or use files from a previous analysis (e.g. last year). This vignette shows only the essential steps needed to make use of `PITcleanr` in preparing data for `DABOM`. 

```{r setup, echo = F}
library(PITcleanr)
library(readr)
library(dplyr)
```


# Minimal R Script

If the user has file containing the complete tag histories for their tags of interest, plus the configuration file and the parent-child file ready, the only script necessary to run in R (which will include saving the output as an Excel workbook) is:

```{r quick-script}
# load the necessary packages
library(PITcleanr)
library(readr)
library(dplyr)

# read in configuration file, config_file = path to that file
my_config = read_csv(config_file)

# read in parent_child file, parent_child_file = path to that file
parent_child = read_csv(parent_child_file) %>%
  addParentChildNodes(configuration = my_config)

# read in PTAGIS observations (ptagis_file), site configuration, and parent-child, 
prepped_df = prepWrapper(ptagis_file = ptagis_file,
                         configuration = my_config,
                         parent_child = parent_child,
                         # filter out detections after a particular date
                         max_obs_date = "20150930",
                         # save results to file
                         save_file = T,
                         file_name = "PITcleanr_output.xlsx")
```

For additional details on site configuration, parent-child tables, and using the `prepWrapper()` function, continue reading.


# Site Configuration

The user may want to start with a configuration file queried from PTAGIS, or generate their own. If they choose to generate their own, it must contain the following columns (at a minimum):

* `site_code`: this is the site code associated with a detection site in PTAGIS
* `config_id`: this is the configuration sequence from PTAGIS associated with a specific antenna arrangement for that site
* `antenna_id`: this is the antenna code or ID from PTAGIS
* `node`: this is the user-defined node that detections are mapped to i.e., how would the user like to group detections on a specific antenna, in a specific configuration, and at a specific site?

The user may want to include other columns such as site name, latitude/longitude, river kilometer (RKM), etc. but none of those are strictly necessary.

The user can start with a template by downloading all the metadata associated with each detection site in PTAGIS using the following:

```{r}
config_template = queryPtagisMeta() %>%
  select(site_code, 
         # site_name,
         config_id = configuration_sequence,
         # antenna_group = antenna_group_name,
         antenna_id) %>%
  # set each node to NA to start
  mutate(node = NA_character_)
```

For a typical DABOM model, where nodes are generally defined as arrays, the user may wish to start with a configuration file that has nodes assigned to each antenna, based on the array the antenna is part of, and merely edit some rows.

```{r, eval = F}
config_template = buildConfig() %>%
  select(site_code, 
         # site_name, 
         config_id, 
         # antenna_group, 
         antenna_id, 
         node)
```


The user may save this template to a .csv file, and then edit it by hand, deleting sites they are not interested in, and filling in the `node` information for all other rows they are interested in.

```{r}
# tell R where to save the csv file
config_file = "configuration.csv"

# save the template as a csv file
write_csv(config_template,
          file = config_file)
```

Once the user has finished editing it, they may read it back into R:

```{r, echo = F}
# find configuration file
config_file = system.file("extdata", 
                          "configuration_TUM.csv", 
                          package = "PITcleanr",
                          mustWork = TRUE)
```

```{r load-config}
# read in configuration file
my_config = read_csv(config_file)
```


# "Parent-Child" Relationships

The parent-child table contains information about how sites or nodes are connected on the stream network. Each site/node child has a single parent, defined as the site/node most directly downstream of the child. Each parent may have multiple children, as the stream network may branch upstream of the parent. For additional information, see that section in [this vignette](Prep_PIT_data.Rmd). 

If the user wishes to create a parent-child table by hand, the file must contain at a minimum the columns `parent` and `child`. This file can then be read into R.

```{r, echo = F}
# find parent-child file
parent_child_file = system.file("extdata", 
                                "parent_child_TUM.csv", 
                                package = "PITcleanr",
                                mustWork = TRUE)
```


```{r load-parent-child}
# read in parent_child file
parent_child = read_csv(parent_child_file)
```

The path to an example parent-child file included with `PITcleanr` can be found by running the following. The user could copy/paste that file to use as a template. Again, only the `parent` and `child` columns are necessary, although other columns may be useful.

```{r, eval = F}
# path to example parent-child table
system.file("extdata", 
            "parent_child_TUM.csv", 
            package = "PITcleanr",
            mustWork = TRUE)
```

The user may choose to only include site codes in the parent and child columns, and then use their configuration file to build out the parent child table to include nodes. To add nodes, use the `addParentChildNodes` function:

```{r add-nodes}
parent_child_nodes = addParentChildNodes(parent_child = parent_child,
                                         configuration = my_config)
```


# Preparing for DABOM

Finally, the user must download complete tag histories from PTAGIS for the tags of interest, ensuring that the following attributes are selected to be included in the output:

* Tag
* Mark Species
* Mark Rear Type
* Event Type
* Event Site Code
* Event Date Time
* Antenna
* Antenna Group Configuration
* Event Release Date Time
* Event Release Site Code

```{r, echo = F}
# find file with PTAGIS observations
ptagis_file = system.file("extdata", 
                          "TUM_Chinook_2015.csv", 
                          package = "PITcleanr",
                          mustWork = TRUE)
```

Now the user can assign those detections to a "node", compress that data, assign directionality and determine which observations should be retained for DABOM, all using the `prepWrapper()` function. The `min_obs_date` argument will filter out observations prior to that date, while the `max_obs_date` argument will ensure that all detections after that date are not retained. 

```{r}
# read in PTAGIS observations and compress them,
# then prep data for choice of which detections to keep
prepped_df = prepWrapper(ptagis_file = ptagis_file,
                         configuration = my_config,
                         parent_child = parent_child_nodes,
                         min_obs_date = "20150301",
                         max_obs_date = "20150930")
```

The user can even save that output directly to an Excel workbook or .csv file by setting `save_file = TRUE` and defining a file path and name in the `file_name` argument. Within this output (either in R, or in the saved Excel file), all the rows where the field `user_keep_obs` is NA or blank must be filled in with either `TRUE` or `FALSE`. The field `auto_keep_obs` provides a suggestion, but the ultimate choice is up to the user. 

```{r, eval = F}
prepped_df = prepWrapper(ptagis_file = ptagis_file,
                         configuration = my_config,
                         parent_child = parent_child_nodes,
                         min_obs_date = "20150301",
                         max_obs_date = "20150930",
                         save_file = T,
                         file_name = "PITcleanr_output.xlsx")
```

#### End Vignette
