---
title: "Reading PIT Tag Data Into R"
author: Kevin See
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Reading PIT Tag Data Into R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
# knitr options
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>"
)

library(knitr)
library(here)
```

# Introduction

Raw PIT tag detections contain a record of every detection of a tag code on an individual antenna, with the associate time-stamp. This data may be queried in primarily two different ways: 1) all of the tag codes detected at a site or antenna, or 2) all the detections of a particular tag code at a variety of sites or antennas. Regardless of what type of query is used, the user is generally dealing with a row of data for every single detection, identified by tag code, site, antenna and date/time (and detections may contain additional data as well).

This data may be stored in a variety of formats or databases. Often these can be broken out into data stored in [PTAGIS](https://ptagis.org/) (The Columbia Basin PIT Tag Information System), and data stored in non-PTAGIS databases (e.g. BioLogic) or local formats (e.g. .xlsx, .txt, .log). One of the first steps in analyzing this data is to read it into R and standardize the format. `PITcleanr` contains a workhorse function, `readCTH()` (for "read complete tag history"), that can accomplish this across many data formats.

# PTAGIS Data

```{r child = "_child_docs/query-ptagis.Rmd"}
```

# Non-PTAGIS Data

```{r child = "_child_docs/query-non-ptagis.Rmd"}
```
