---
output:
  rmarkdown::html_vignette: default
  html_document: default
---
```{r, echo = FALSE, message = FALSE, warning = FALSE, results = "hide"}
# knitr options
knitr::opts_chunk$set(
collapse = TRUE,
warning = FALSE,
message = FALSE,
echo = TRUE,
comment = "#>"
)

library(knitr)
```

The Columbia Basin PIT Tag Information System ([PTAGIS](https://ptagis.org/)) is the centralized regional database for PIT-tag detections within the Columbia River Basin. It contains a record of each detection of every PIT tag, including the initial detection, or "mark", when the tag is embedded in the fish, detections on PIT-tag antennas, recaptures (e.g. at weirs) and recoveries (e.g. carcass surveys). It contains a record of every individual detection, which means potentially multiple records of a tag being detected on the same antenna over and over e.g., in the case that it is not moving. Therefore, querying PTAGIS for all of these detections leads to a wealth of data, which can be unwieldy for the user. `PITcleanr` aims to compress that data to a more manageable size, without losing any of the information contained in that dataset. 

`PITcleanr` starts with a complete capture history query from [PTAGIS](https://ptagis.org/) for a select group of tags of interest. The user will need to compile this list of tags themselves, ideally in a .txt file with one row per tag number, to make it easy to upload to a PTAGIS query. 

For convenience, we've included one such file with `PITcleanr`. The following code can be used to find the path to that example file, titled "tag_list_TUM_2015.txt", containing IDs for Chinook salmon adults implanted with PIT tags at Tumwater Dam in 2015. Alternatively, the user can follow along with their own file containing their tags of interest.

```{r tag-list-file, eval = F}
system.file("extdata", 
            "tag_list_TUM_2015.txt", 
            package = "PITcleanr",
            mustWork = TRUE)
```

The file of tag codes is very simple:

```{r, echo = F}
system.file("extdata", 
            "tag_list_TUM_2015.txt", 
            package = "PITcleanr",
            mustWork = TRUE) %>%
  readr::read_table(col_names = F)
```

Once the user has created their own tag list, or located this example one, they can go to the [PTAGIS homepage](https:://www.ptagis/org). From the homepage, go to the [Advanced Reporting](https://www.ptagis.org/data/advanced-reporting) page, which can also be found under the Data tab on the homepage. To access Advanced Reporting, the user will need a free account from PTAGIS, and to be logged in. Once on the Advanced Reporting page, select "Launch" and create a custom query by selecting "Create Query Builder2 Report", and choosing the "Complete Tag History" query type.  

There are several query indices on the left side of the query builder, but for the purposes of `PITcleanr` only a few are needed. First, under "1 Select Attributes" the following fields are required to work with `PITcleanr`:

* Tag
* Mark Species
* Mark Rear Type
* Event Type
* Event Site Type
* Event Site Code
* Event Date Time
* Antenna
* Antenna Group Configuration
* Event Release Site Code
* Event Release Date Time

Simply move these fields over to the "Selected:" column on the page. Other fields may be included as well, but the ones listed above must be added. Any additional fields will just be included as extra columns in the query output. 

The only other required index is "2 Select Metrics", but that can remain as the default, "CTH Count", which provides one record for each event recorded per tag.

Set up a filter for specific tags by next navigating to the "29 Tag Code - List or Text File" on the left. And then, after selecting "Tag" under "Attributes:", click on "Import file...". Simply upload the .txt file containing your PIT tag codes of interest, or alternatively, feel free to use the "tag_list_TUM_2015.txt" file provided with `PITcleanr`. After choosing the file, click on "Import" and the tag list will be loaded (delimited by semi-colons). Click "OK". 

Under "Report Message Name:" near the bottom, name the query something appropriate, such as "TUM_Chinook_2015", and select "Run Report". Once the query has successfully completed, the output can be exported as a .csv file (e.g. "TUM_Chinook_2015.csv"). Simply click on the "Export" icon near the top, which will open a new page, and select the default settings:

* Export: Whole report
* CSV file format
* Export Report Title: unchecked
* Export filter details: unchecked
* Remove extra column: Yes

And click "Export", again. We have included an example of data downloaded from PTAGIS with the `PITcleanr` package. The user can find where that example file has been saved by running the following code. The file path will be stored as a new object `ptagis_file`.

```{r}
ptagis_file = system.file("extdata", 
                          "TUM_Chinook_2015.csv", 
                          package = "PITcleanr",
                          mustWork = TRUE)
```

Alternatively, if the user has run a query from PTAGIS as described above, they should set `ptagis_file` to the path and file name of the .csv they downloaded.

```{r, eval = F}
# As an example, set path to PTAGIS query output
ptagis_file = "C:/Users/usernamehere/Downloads/TUM_Chinook_2015.csv"

```

```{r raw-ptagis-data, echo = F}
raw_ptagis = readr::read_csv(ptagis_file)
n_raw_tags = n_distinct(raw_ptagis$`Tag Code`)

mark_only_tags = raw_ptagis %>%
  select(`Tag Code`, `Event Type Name`) %>%
  count(`Tag Code`, `Event Type Name`) %>%
  pivot_wider(names_from = `Event Type Name`,
              values_from = n,
              values_fill = 0) %>%
  #mutate(nDetections = rowSums(across(where(is.numeric)))) %>%
  filter(Observation == 0 & Recapture == 0 & Recovery == 0)

```

Note that in our example file, there are `r nrow(raw_ptagis)` detections (rows) for `r n_raw_tags` unique tags, matching the number of tags in our example tag list "tag_list_TUM_2015.txt". For a handful of those tags, in our case `r nrow(mark_only_tags)`, there is only a "Mark" detection i.e., that tag was never detected again after the fish was tagged and released from Tumwater Dam. For the remaining, there are many tags often detected at the same site and sometimes on the same antenna. Data like this, while full of information, can be difficult to analyze efficiently. To illustrate, here is an example of the raw data for a single tag:

```{r, echo = F}
raw_ptagis %>%
  select(`Tag Code`) %>%
  distinct() %>%
  # slice(1:3) %>%
  slice(2) %>%
  left_join(raw_ptagis) %>%
  kable()
  # DT::datatable()

```

Let's "compress" all of those detections into a more manageable format as follows...
