---
output:
  rmarkdown::html_vignette: default
  html_document: default
---

The Columbia Basin PIT Tag Information System ([PTAGIS](https://ptagis.org/)) is the centralized regional database for PIT-tag detections within the Columbia River Basin. It contains a record of each detection of every PIT tag, including the initial detection, or "mark", when the tag is implanted in the fish, detections on PIT-tag antennas, recaptures (e.g. at weirs) and recoveries (e.g. carcass surveys). It contains a record of every individual detection, which means potentially multiple records of a tag being detected on the same antenna over and over e.g., in the case that it is not moving. Therefore, querying PTAGIS for all of these detections leads to a wealth of data, which can be unwieldy for the user. `PITcleanr` aims to compress that data to a more manageable size, without losing any of the information contained in that dataset. 

`PITcleanr` starts with a complete capture history query from [PTAGIS](https://ptagis.org/) for a select group of tags of interest. The user will need to compile this list of tags themselves, ideally in a .txt file with one row per tag number, to make it easy to upload to a PTAGIS query. 

For convenience, we've included one such file with `PITcleanr`. The following code can be used to find the path to that example file that was installed with the package, titled "tag_list_TUM_2015.txt", containing IDs for Chinook salmon adults implanted with PIT tags at Tumwater Dam in 2015. Feel free to copy/paste that file to use as a template for creating your own tag list, in which case, you can follow along with your own tags of interest.

```{r tag-list-file, eval = F}
system.file("extdata", 
            "tag_list_TUM_2015.txt", 
            package = "PITcleanr",
            mustWork = TRUE)
```

The example file of tag codes is very simple:

```{r, echo = F}
system.file("extdata", 
            "tag_list_TUM_2015.txt", 
            package = "PITcleanr",
            mustWork = TRUE) %>%
  readr::read_table(col_names = F)
```

Once the user has created their own tag list, or located this example one, they can go to the [PTAGIS homepage](https:://www.ptagis/org) to query the complete tag histories for those tags. The complete tag history query is available to under [Advanced Reporting](https://www.ptagis.org/Data/AdvancedReporting), which requires a free account from PTAGIS. From the [homepage](https:://www.ptagis/org), click on "Login/Register", and either login to your existing account, or click "Create a new account" to create one. 

Once logged into your account, scroll down the dashboard page to the Advanced Reporting Links section. PTAGIS allows users to save reports/queries to be run again. If you've created one for PITcleanr in the past, and saved it, you can locate it by clicking on the "My Reports" link. If you'd like to create a new query, click on "Query Builder", or "Advanced Reporting Home Page" and then ""Create Query Builder2 Report". From here, choose "Complete Tag History" from the list of possible reports.

There are several query indices on the left side of the query builder, but for the purposes of `PITcleanr` only a few are needed. First, under "1 Select Attributes" the following attributes are required to work with `PITcleanr`:

* Tag
* Event Site Code
* Event Date Time
* Antenna
* Antenna Group Configuration

This next group of attributes are not required, but are highly recommended:

* Mark Species
* Mark Rear Type
* Event Type
* Event Site Type
* Event Release Site Code
* Event Release Date Time


Simply move these attributes over from the "Available" column to the "Selected:" column on the page. Other fields of interest to the user may be included as well (e.g. Event Length), but the ones listed above must be added. Any additional attributes will just be included as extra columns in the query output. 

The only other required index is "2 Select Metrics", but that can remain as the default, "CTH Count", which provides one record for each event recorded per tag.

Set up a filter for specific tags by next navigating to the "29 Tag Code - List or Text File" on the left. And then, after selecting "Tag" under "Attributes:", click on "Import file...". Simply upload the .txt file containing your PIT tag codes of interest, or alternatively, feel free to use the "tag_list_TUM_2015.txt" file provided with `PITcleanr`. After choosing the file, click on "Import" and the tag list will be loaded (delimited by semi-colons). Click "OK". 

Under "Report Message Name:" near the bottom, name the query something appropriate, such as "TUM_Chinook_2015", and select "Run Report". Once the query has successfully completed, the output can be exported as a .csv file (e.g. "TUM_Chinook_2015.csv"). Simply click on the "Export" icon near the top, which will open a new page, and select the default settings:

* Export: Whole report
* CSV file format
* Export Report Title: unchecked
* Export filter details: unchecked
* Remove extra column: Yes

And click "Export", again. We have included an example of data downloaded from PTAGIS with the `PITcleanr` package. The user can identify where that example file has been saved by running the following code. The file path will be stored as a new object `ptagis_file`.

```{r}
ptagis_file = system.file("extdata", 
                          "TUM_Chinook_2015.csv", 
                          package = "PITcleanr",
                          mustWork = TRUE)
```

Alternatively, if the user has run a query from PTAGIS as described above, they could set `ptagis_file` to the path and file name of the .csv they downloaded.

```{r, eval = F}
# As an example, set path to PTAGIS query output
ptagis_file = "C:/Users/usernamehere/Downloads/TUM_Chinook_2015.csv"

```

```{r raw-ptagis-data, echo = F}
raw_ptagis = readr::read_csv(ptagis_file,
                             show_col_types = F)
n_raw_tags = n_distinct(raw_ptagis$`Tag Code`)

mark_only_tags = raw_ptagis %>%
  select(`Tag Code`, `Event Type Name`) %>%
  count(`Tag Code`, `Event Type Name`) %>%
  pivot_wider(names_from = `Event Type Name`,
              values_from = n,
              values_fill = 0) %>%
  #mutate(nDetections = rowSums(across(where(is.numeric)))) %>%
  filter(Observation == 0 & Recapture == 0 & Recovery == 0)

```

Note that in our example file, there are `r nrow(raw_ptagis)` detections (rows) for `r n_raw_tags` unique tags, matching the number of tags in our example tag list "tag_list_TUM_2015.txt". For a handful of those tags, in our case `r nrow(mark_only_tags)`, there is only a "Mark" detection i.e., that tag was never detected again after the fish was tagged and released. For the remaining tags, many of them were often detected at the same site and sometimes on the same antenna. Data like this, while full of information, can be difficult to analyze efficiently. To illustrate, here is an example of some of the raw data for a single tag:

```{r, echo = F}
raw_ptagis %>%
  select(`Tag Code`) %>%
  distinct() %>%
  slice(2) %>%
  left_join(raw_ptagis,
            by = join_by(`Tag Code`),
            multiple = "all") %>%
  slice(c(1:15)) %>%
  kable() %>%
  kableExtra::kable_styling()
```
