---
output:
  rmarkdown::html_vignette: default
  html_document: default
---

`PITcleanr` also includes a function `buildConfig()` that maps each array (row of antennas) at a site onto a node. Using `buildConfig()`, the upstream arrays are labeled with the site code plus `A0` and the downstream arrays are assigned `B0`. For sites with three arrays, the middle array is grouped with the upstream array. For sites with four arrays, the upper two arrays are mapped to `A0` and the lower two arrays are mapped to `B0`. Alternatively, the user could use [metadata from PTAGIS](https://ptagis.org/sites) (also see `?queryPtagisMeta`) and construct this mapping any way they choose. This type of configuration file (or tibble or data frame) can be provided to the `configuration` argument in `compress()` to assign each detection to it's associated node.

In the following example, we use the `buildConfig()` function to generate a default configuration and save it as `array_configuration`. Several sites are then consolidated into a single node (e.g. LNF, TUM), and some mark-recovery-recapture sites are merged with upstream array nodes, and the modified configuration is saved as `configuration` which can then be fed to the `compress()` function.

```{r}
array_configuration = buildConfig()

# customize some pieces
configuration = array_configuration %>%
  # first, for example, 'LNF' and 'LEAV' are re-coded into a single node 'LNF'
  mutate(node = ifelse(site_code %in% c('LNF', 'LEAV'),
                       'LNF',
                       node),
         # these three nodes are all re-coded to a single 'TUM'
         node = ifelse(site_code %in% c('TUF', 'TUMFBY', 'TUM'),
                       'TUM',
                       node),
         node = ifelse(site_code == 'CHIWAC',
                       'CHWA0',
                       node),
         node = ifelse(site_code %in% c('CHIWAR', 'CHIWAT'),
                       'CHLA0',
                       node),
         node = ifelse(site_code == 'CHIW',
                       'CHLA0',
                       node),
         # In this case, PIT tags from carcass recoveries in the Chikamin River are
         # grouped with the upper 'CHU' array
         node = ifelse(site_code == 'CHIKAC',
                       'CHUA0',
                       node),
         node = ifelse(site_code == 'NASONC',
                       'NALA0',
                       node),
         node = ifelse(site_code == 'WHITER',
                       'WTLA0',
                       node),
         node = ifelse(site_code == 'LWENAT',
                       'LWNA0',
                       node)) %>%
  distinct()
```

Now, re-run the `compress()` function, except supplying a `configuration`:

```{r}
# re-run compress(), providing configuration
comp_obs = compress(ptagis_file,
                    configuration = configuration)

# look at first part of comp_obs
head(comp_obs)
```
