---
output:
  rmarkdown::html_vignette: default
  html_document: default
---

The `PITcleanr` package can be installed as an R package from GitHub by using Hadley Wickham's `devtools` package:

```{r gh-installation, eval = FALSE}
# install and load remotes, if necessary
install.packages("devtools")
remotes::install_github("KevinSee/PITcleanr", 
                         build_vignettes = TRUE)
```

`devtools` may require a working development environment. Further details around the installation of `devtools` can be found [here](https://www.r-project.org/nosvn/pandoc/devtools.html).

* For Windows, that will involve the downloading and installation of Rtools. The latest version of Rtools can be found [here](https://cran.r-project.org/bin/windows/Rtools/). 
* For Mac OS, that may involve installing Xcode from teh Mac App Store.
* For Linux, that may involve installing a compiler and various development libraries, depending on the specific version of Linux.

To install the latest development version of `PITcleanr`:

```{r develop-version, eval = FALSE}
remotes::install_github("KevinSee/PITcleanr@develop")
```

Alternatively, the `PITcleanr` compendium can be downloaded as a zip file from from this URL: <https://github.com/KevinSee/PITcleanr/archive/master.zip>
Once extracted, the functions can be sourced individually, or a user can build the R package locally.
