---
output:
  rmarkdown::html_vignette: default
  html_document: default
---

The complete tag history query output from [PTAGIS](https://ptagis.org/) (e.g., `ptagis_file`) will provide a record for every detection of each tag code in the tag list. Again, this may include multiple detections on the same antenna, or the same site within a short period of time, leading to an unwieldy and perhaps messy dataset. One of the main purposes of `PITcleanr` is to compress that data into an initial summary using the function `compress()`. At a minimum, this function requires a path e.g., `ptagis_file`, to the complete tag history query results downloaded from PTAGIS. 

In our example, we perform the `compress()` function on the `ptagis_file` object containing the file path to our query results, and write the output to an object `comp_obs` containing the compressed observations.

```{r, eval = T}
# view path to example file, of course you can also set ptagis_file to your own PTAGIS query results
ptagis_file

# run compress() function on it
comp_obs = compress(ptagis_file)

# look at first parts of resulting object
head(comp_obs)

```

The output consists of a tibble containing columns for:

* **tag_code:** The unique PIT tag ID.
* **node:** By default, each site code from PTAGIS is considered a node. More on this below... 
* **slot:** A detection "slot" for each tag, numbered in chronological order. Also more on this below... 
* **event_type_name:** The type of "event". Typically, mark, observation, recapture, or recovery.
* **n_dets:** The number of detections that occurred within that slot.
* **min_det:** The time of the first (min) detection in the slot.
* **max_det:** The time of the last (max) detection in the slot.
* **duration:** The duration of that slot (maximum - minimum detection time). 
* **travel_time:** The travel time between the previous slot and that one. 

***A note on "nodes"***: A node is the spatial scale of interest for the user. By default, the `compress()` function considers a site code from PTAGIS as a node. However, a node could be defined as the individual PIT antenna a detection was made on, or the array that antenna is a part of, or groups of arrays, or sites, or groups of sites, or possibly even larger (e.g, any detection in **this** tributary!) depending on the spatial scale desired. The user may decide to define some arrays at particular sites to be their own nodes, while simultaneously lumping all the sites in a particular watershed into a single node. To utilize this kind of grouping, a configuration file or table must be supplied to the `configuration` argument in the `compress()` function. This configuration file helps map what node each antenna or MRR site corresponds to. By default, `compress` assigns each site code to its own node.

Each slot in the output is defined as all detections on a particular node before the tag is detected on a different node. The user can define a maximum number of minutes between detections before a new slot should be defined by supplying a value to the `max_minutes` argument to `compress()`. The units of the duration and travel_time columns can also be defined by the `units` argument. The default is minutes (`mins`). As an example, if a tag moves from node A to B and back to A, there will be three slots in the compressed data.

The help menu for `compress()`, or any function for that matter, can be accessed using:
```{r, eval = F}
?PITcleanr::compress

```
